Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-04 Paper_Draft2/Draft2.Rmd

author:
  - name: Amrita Acharya
    affil: 1
  - name: Dianne Caravela
    affil: 1
  - name: Eunice Kim
    affil: 1
  - name: Emma Kornberg
    affil: 1
  - name: Elisabeth Nesmith
    affil: 1
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
    email: bbaumer@smith.edu

journal: water
type: article
status: submit
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
#appendix: appendix.tex
classoption: fleqn
correspondence: |
  aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu

@@

The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we used to debias the COMPAS algorithm [@aif360-oct-2018]. Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally. An algorithm has no understanding of the historical oppression of certain groups and how such bias is baked into the data on which the algorithm is trained. Thus, fairness metrics provide a method of evaluating an algorithm’s level of bias towards or against the unprivileged group versus the privileged group. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project. To begin, we implemented a logistic regression model predicting recidivism using the defendant data. This analyzes the baseline values without any bias mitigation. We use logistic regression, as it is the easiest model to interpret in the given context.

Fairness is subjective; one person might consider an algorithm fair if groups are given the same treatment, while someone else might only consider the algorithm fair if groups as a whole receive the same outcomes. In our research we choose to use the definition of group fairness as our definition for assessing fairness in the processing approach. Group fairness metrics take into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent systemic issues. In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021]. Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse [@kypraiou_what_2021]. We chose the following four group fairness metrics to evaluate our models.

@@

The prejudice remover is a method for reducing indirect prejudice (i.e., how COMPAS is racially biased because it uses proxy variables for race). The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification [@kamishima2012fairness]. The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by @kamishima2012fairness. This in turn enforces a classifier’s independence from sensitive information (e.g., race). Similarly to pre-processing, we compared the results of our in-processing methods with both the baseline and the pre-processing model to gauge which method so far has the better group fairness.

@@

This document contains 4793 words

>>>>

author:
  - name: Amrita Acharya
    affil: 1
  - name: Dianne Caravela
    affil: 1
  - name: Eunice Kim
    affil: 1
  - name: Emma Kornberg
    affil: 1
  - name: Elisabeth Nesmith
    affil: 1
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
    email: bbaumer@smith.edu
#journal: water
#status: submit
type: article
bibliography: mybibfile.bib
csl: ../resources/chicago-author-date.csl
#appendix: appendix.tex
classoption: fleqn
correspondence: |
  aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu

@@

The AIF360 toolkit contains various group and individual fairness metrics as well as pre-processing, in-processing, and post-processing algorithms that we use to debias the COMPAS algorithm [@aif360-oct-2018]. Fairness metrics are mathematical measures of whether an algorithm treats members of different groups (such as racial or gender groups) equally. An algorithm has no understanding of the historical oppression of certain groups and how such bias is baked into the data on which the algorithm is trained. Thus, fairness metrics provide a method of evaluating an algorithm’s level of bias towards or against the unprivileged group versus the privileged group. We researched the definitions and applications of different fairness metrics [@ashokan2021fairness] to determine which metric would be most appropriate for our project. To begin, we implemented a logistic regression model predicting recidivism using the defendant data. This analyzes the baseline values without any bias mitigation. We use logistic regression, as it is the easiest model to interpret in the given context.

Fairness is subjective; one person might consider an algorithm fair if groups are given the same treatment, while someone else might only consider the algorithm fair if groups as a whole receive the same outcomes. In our research we choose to use the definition of group fairness as our definition for assessing fairness in the processing approaches. Group fairness metrics take into account the attributes of a whole group as opposed to just one individual in the group, allowing us to represent systemic issues. In general, group fairness metrics require that the unprivileged group is treated similarly to the privileged group, whereas individual fairness metrics require individuals to be treated consistently [@kypraiou_what_2021]. Group and individual metrics work in opposition of one another, meaning that when group fairness improves, individual fairness gets worse [@kypraiou_what_2021]. We chose the following four group fairness metrics to evaluate our models.

@@

The prejudice remover is a method for reducing indirect prejudice (i.e., how COMPAS is racially biased because it uses proxy variables for race). The prejudice remover implements two different regularizers, one to avoid overfitting and one to enforce fair classification [@kamishima2012fairness]. The prejudice remover regularizer works by minimizing the prejudice index, a mathematical equation for quantifying fairness defined by @kamishima2012fairness. This in turn enforces a classifier’s independence from sensitive information (e.g., race). Similar to pre-processing, we compare the results of our in-processing methods with both the baseline and the pre-processing model to gauge which method so far has the better group fairness.

@@

```{r word count, include=F}
wordcountaddin:::text_stats()
```

### 2022-05-04 Paper_Draft2/Draft2.tex

%=================================================================
\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}

@@

Our project builds on Women at the Table's various debiasing algorithms
used in its AI \& Equality Human Rights Toolbox to conduct our own
analyses on the COMPAS data set. Based on this analysis, we employ a
human rights framework to contribute to the ProPublica and Northpointe
debate and investigate whether or to what extent there is racial bias in
the COMPAS algorithm. With a solid understanding of the two sides, we
aim to pinpoint the shortcomings of both arguments and correct them in
our analyses. We will use various debiasing techniques and fairness
metrics to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this library addition is the COMPAS General
Recidivism Risk Scores data set from the
\href{https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html\#aif360.datasets.CompasDataset}{AI
Fairness 360} (AIF360) toolkit. The AIF360 toolkit builds on the dataset
released by ProPublica collaborators Larson et al.~created to examine
the racial bias and the true outcomes of the recidivism risk scores in
the COMPAS algorithm for the initial
\href{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}{``Machine
Bias''} article. For this, Larson et al.~obtained two years worth of
COMPAS scores from the Broward County Sheriff's Office in Florida, as
well as the corresponding intake information for each defendant
including but not limited to name, sex, race, age, and charge degree and
description. They also obtained data about whether defendants actually
recidivated or not in the two year period following their initial COMPAS
score assessment. AIF360 then processed the data with the same
procedures that Larson et al.~followed for their analysis.

@@

\caption{A snippet of the data set we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
\end{figure}

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied \citet{angwin2016machine}.

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used to debias the COMPAS algorithm
\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
whether an algorithm treats members of different groups (such as racial
or gender groups) equally. An algorithm has no understanding of the
historical oppression of certain groups and how such bias is baked into
the data on which the algorithm is trained. Thus, fairness metrics
provide a method of evaluating an algorithm's level of bias towards or
against the unprivileged group versus the privileged group. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. To begin, we implemented a logistic
regression model predicting recidivism using the defendant data. This
analyzes the baseline values without any bias mitigation. We use
logistic regression, as it is the easiest model to interpret in the
given context.

Fairness is subjective; one person might consider an algorithm fair if
groups are given the same treatment, while someone else might only
consider the algorithm fair if groups as a whole receive the same
outcomes. In our research we choose to use the definition of group
fairness as our definition for assessing fairness in the processing
approach. Group fairness metrics take into account the attributes of a
whole group as opposed to just one individual in the group, allowing us
to represent systemic issues. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse
\citep{kypraiou_what_2021}. We chose the following four group fairness
metrics to evaluate our models.

@@

This metric is best used when it is very important to catch positive
outcomes while false positives are not exceptionally problematic
\citep{Cortez2019How}. This is not the case for the COMPAS data set, as
false positives mean extra jail time for someone who will not actually
re-offend.

@@

The prejudice remover is a method for reducing indirect prejudice (i.e.,
how COMPAS is racially biased because it uses proxy variables for race).
The prejudice remover implements two different regularizers, one to
avoid overfitting and one to enforce fair classification
\citep{kamishima2012fairness}. The prejudice remover regularizer works
by minimizing the prejudice index, a mathematical equation for
quantifying fairness defined by \citet{kamishima2012fairness}. This in
turn enforces a classifier's independence from sensitive information
(e.g., race). Similarly to pre-processing, we compared the results of
our in-processing methods with both the baseline and the pre-processing
model to gauge which method so far has the better group fairness.

>>>>

%=================================================================
\documentclass[,article,,moreauthors,pdftex]{mdpi}

@@

Our project builds on Women at the Table's various debiasing algorithms
used in its AI \& Equality Human Rights Toolbox to conduct our own
analyses on the COMPAS dataset. Based on this analysis, we employ a
human rights framework to contribute to the ProPublica and Northpointe
debate and investigate whether or to what extent there is racial bias in
the COMPAS algorithm. With a solid understanding of the two sides, we
aim to pinpoint the shortcomings of both arguments and correct them in
our analyses. We will use various debiasing techniques and fairness
metrics to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this library addition is the COMPAS General
Recidivism Risk Scores dataset from the
\href{https://aif360.readthedocs.io/en/stable/modules/generated/aif360.datasets.CompasDataset.html\#aif360.datasets.CompasDataset}{AI
Fairness 360} (AIF360) toolkit. The AIF360 toolkit builds on the dataset
released by ProPublica collaborators Larson et al.~created to examine
the racial bias and the true outcomes of the recidivism risk scores in
the COMPAS algorithm for the initial
\href{https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing}{``Machine
Bias''} article. For this, Larson et al.~obtained two years worth of
COMPAS scores from the Broward County Sheriff's Office in Florida, as
well as the corresponding intake information for each defendant
including but not limited to name, sex, race, age, and charge degree and
description. They also obtained data about whether defendants actually
recidivated or not in the two year period following their initial COMPAS
score assessment. AIF360 then processed the data with the same
procedures that Larson et al.~followed for their analysis.

@@

\caption{A snippet of the dataset we will be using, containing information on a defendant's age, sex, race, criminal history, charge degree, charge description, and two-year recidivism outcome.}\label{fig:table snip}
\end{figure}

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple datasets, including the
COMPAS dataset that accompanied \citet{angwin2016machine}.

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we use to debias the COMPAS algorithm
\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
whether an algorithm treats members of different groups (such as racial
or gender groups) equally. An algorithm has no understanding of the
historical oppression of certain groups and how such bias is baked into
the data on which the algorithm is trained. Thus, fairness metrics
provide a method of evaluating an algorithm's level of bias towards or
against the unprivileged group versus the privileged group. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. To begin, we implemented a logistic
regression model predicting recidivism using the defendant data. This
analyzes the baseline values without any bias mitigation. We use
logistic regression, as it is the easiest model to interpret in the
given context.

Fairness is subjective; one person might consider an algorithm fair if
groups are given the same treatment, while someone else might only
consider the algorithm fair if groups as a whole receive the same
outcomes. In our research we choose to use the definition of group
fairness as our definition for assessing fairness in the processing
approaches. Group fairness metrics take into account the attributes of a
whole group as opposed to just one individual in the group, allowing us
to represent systemic issues. In general, group fairness metrics require
that the unprivileged group is treated similarly to the privileged
group, whereas individual fairness metrics require individuals to be
treated consistently \citep{kypraiou_what_2021}. Group and individual
metrics work in opposition of one another, meaning that when group
fairness improves, individual fairness gets worse
\citep{kypraiou_what_2021}. We chose the following four group fairness
metrics to evaluate our models.

@@

This metric is best used when it is very important to catch positive
outcomes while false positives are not exceptionally problematic
\citep{Cortez2019How}. This is not the case for the COMPAS dataset, as
false positives mean extra jail time for someone who will not actually
re-offend.

@@

The prejudice remover is a method for reducing indirect prejudice (i.e.,
how COMPAS is racially biased because it uses proxy variables for race).
The prejudice remover implements two different regularizers, one to
avoid overfitting and one to enforce fair classification
\citep{kamishima2012fairness}. The prejudice remover regularizer works
by minimizing the prejudice index, a mathematical equation for
quantifying fairness defined by \citet{kamishima2012fairness}. This in
turn enforces a classifier's independence from sensitive information
(e.g., race). Similar to pre-processing, we compare the results of our
in-processing methods with both the baseline and the pre-processing
model to gauge which method so far has the better group fairness.
